A comparison of single-cell trajectory inference methods: towards more accurate and robust tools

### 2018-10-01 manuscript/paper.Rmd
update paper with scalability assessment

In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←. Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method. Instead, the choice toof method →mostly← depends on the trajectory topology present in the data. →To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.

@@

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`). →Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the prior information they require and the outputs they produce. Two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies can be detected←. Early TI methods typically fixed the topology algorithmically (such as linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along the fixed topology. →Most recent methods also infer the topology [@street_slingshotcelllineage_2017; @wolf_graphabstractionreconciles_2017], which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←

@@

# Results
## Method overview
<!-- Method characterisation -->
In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")`, `r ref("sfig", "overview_evaluation_expanded")`). We included methods primarily based on their free availability and presence of a programming interface (`r ref("stable", "tools")`). In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. The output of each method was converted to this model using different wrapper types (`r ref("fig", "results_summary", "a")`).

When required, we also provided prior information to the method, which ranges from "soft" priors, such as start or end cells, to "harder" priors such as the known grouping of the cells (`r ref("fig", "results_summary", "a")`).

The main difference between trajectory inference methods is whether the method fixes the topology, and if it is free, what kind of topologies it can detect. We defined nine possible types of topology, ranging from the very basic (linear, cyclical and bifurcating) to the most complex (connected and disconnected graphs). Most methods either focus on inferring linear trajectories, or limit the search to tree or less complex topologies, with only a selected few attempting to infer cyclic or disconnected topologies (`r ref("fig", "results_summary", "a")`).

<!-- Overall evaluation results -->
Each method was evaluated on four core aspects: (i) benchmarking given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) their scalability with both the number of cells and features (iii) the stability of the predictions across multiple bootstrap replicates, and (iv) the user friendliness of the provided software and documentation. Overall, we found a large diversity across all four evaluation criteria, with only a few methods, such as PAGA, GNG, Slingshot, Projected TSCAN and SCORPIUS, performing well across the board  (`r ref("fig", "results_summary", "b")`). We will first discuss each evaluation criterion in detail, after which we conclude with guidelines for method users and future perspectives for method developers.
```{r}
add_fig(
  fig_path = result_file("results_summary.pdf", experiment_id = "08-summary"),
  ref_id = "results_summary", 
  caption_main = glue::glue("A characterisation of the {n_methods_evaluated} methods evaluated in this study, and their overall evaluation results."), 
  caption_text = "(a) We characterised the methods according to their required priors, the types of inferreable topologies, whether or not the inferred topology is constrained by the algorithm or a parameter. The methods are grouped vertically based on the most complex trajectory type they can infer. (b) Overall results of the evaluation on four criteria: the benchmarking using a reference trajectory on real and synthetic data, the scalability with increasing number of cells and features, the stability across dataset bootstraps, and the quality control of the tool which implemented the method. ",
  width = 15,
  height = 18
)
```

```{r}
add_fig(
  fig_path = result_file("results_detailed.pdf", experiment_id = "08-summary"),
  ref_id = "results_detailed", 
  caption_main = glue::glue("Detailed results of the four main evaluation criteria: benchmarking, scalability, stability and quality control."), 
  caption_text = "(a) Results of the benchmark across metrics, dataset sources and dataset trajectory types. The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type. Also shown is how often a method errored, and what the reasons were for these errors. (b) Scalability results, with increasing number of cells and features. Shown are predicted times at a fixed set of cells (10.000) or features (10.000). (c) Stability results (d) Quality control results across six categories.",
  width = 15,
  height = 18
)
```
## Benchmark
<!-- Benchmark methods -->
We defined several possible metrics to compare a prediction to a reference trajectory (`r ref("snote", "metrics")`). Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the topology (`r label_metric("him")`), the quality of clustering of the branch assignment  (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`)(`r ref("fig", "ti_evaluation_overview", "d")`).The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).  Some real datasets were classified as gold if the reference trajectory was defined using other data than the expression data, such as via cellular sorting. All other real datasets were classified as silver. For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011]. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.

<!-- Benchmark overview -->
We found that method performance was very variable across datasets, indicating that there is no "one-size-fits-all" method that works well on every dataset `r ref("sfig", "benchmark_interpretation", "a")`. Even methods which can detect most of the trajectory types, such as PAGA or GNG, were not the best methods across all trajectory types `r ref("fig", "results_detailed", "a")`. The average performance between the different dataset sources was moderately to highly similar correlated, with the performance on all dataset sources being moderately to highly correlated (0.5 - 0.9) with the one on real gold datasets `r ref("sfig", "benchmark_interpretation", "b")`. On the other hand, the different metrics frequently disagreed with each other, with Monocle scoring better on the topology scores, while other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches `r ref("fig", "results_detailed", "a")`.

<!-- Topologies-->
The performance of a method was strongly dependent on the type of trajectory present in the data `r ref("fig", "results_detailed", "a")`. Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories. This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, while those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (`r ref("sfig", "benchmark_interpretation", "c")`).

<!-- Complementarity between methods -->
The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods. To test this, we looked for subsets of methods at which the probability of getting a "top" model (defined as having a score at least 95% as the best model) was highest. On data which contains a tree trajectory, the use of the best method, RaceID / StemID, would result in getting top model about 50% of the time. This probability rapidly increased to up to 90% with the addition of 2 to 5 other methods (`r ref("fig", "complementarity", "a")`). We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`). When the datasets from different trajectory types were pooled, the number of methods necessary to achieve more than 90% top models increased to more than 8. This pooling also resulted in a more diverse set of methods, containing both strictly linear methods such as SCORPIUS, and methods with a broad trajectory type range such as PAGA. Altogether, this shows that there is considerably complementarity between the different methods, and that users should try out different methods on their data.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = result_file("complementarity.rds", experiment_id = "10-benchmark_interpretation"),
  ref_id = "complementarity", 
  caption_main = "Complementarity between different TI methods", 
  caption_text = "",
  width = 12,
  height = 6
)
```

```{r echo=FALSE, results="asis"}
add_sfig(
  fig_path = result_file("benchmark_interpretation.rds", experiment_id = "10-benchmark_interpretation"),
  ref_id = "benchmark_interpretation", 
  caption_main = "Robustness of the benchmark", 
  caption_text = "Shown are the distributions of the difference in size between predicted and reference topologies. A positive difference means that the topology predicted by the method is more complex than the one in the reference.",
  width = 12,
  height = 15
)
```

## Scalability
Given the ever increasing 
## Stability
←
## Quality control

@@

We observed no clear relation between method quality and method performance (`r ref("fig", "results_summary")`). We could also not find any quality aspect which was significantly associated with method performance.

@@

<!-- FOR METHOD USERS -->
<!-- Practical guidelines -->
Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")` and `r print_url("guidelines.dynverse.org")`). We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data. For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure. In each of these use cases, →our evaluation suggests a different set of optimal methods, as shown in `r ref("fig", "user_guidelines")`. Several other factors will also impact the choice of methods, such as the dimensions of the dataset, and the prior information which is available. These factors and several others can all be changed in our interactive app (`r print_url("http://dynverse.org")`). This app can also be used to explore the results of this evaluation, such as filtering the datasets, or changing the importance of the evaluation metrics for the final ranking.←

@@

### Aggregation of scores
To rank methods, we needed to aggregate the different scores on two levels: across datasets and across different metrics. This aggregation strategy is explained into more detail in `r ref("snote", "metrics")`.

@@

Next, we calculated for each metric, trajectory type and dataset source an average score using an arithmetic mean.

Next, we aggregated the scores across trajectory types and dataset source (real/gold, real/silver, synthetic/dyngen, synthetic/splatter, synthetic/prosstt and synthetic/dyntoy), so that each of these had an equal weight on the final score:

1. Calculate a score for each method, dataset source and trajectory type
←
## Benchmark
### Method execution
Each execution of a method on a dataset was performed in a separate task as part of a gridengine job. Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 @ 2.40GHz, and one R session was started for each task. During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution. If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”. The environment variable `R_MAX_NUM_DLLS` set to 500. The `base::set.seed` was overridden in order to prevent stochastic TI methods from pretending to be deterministic and/or robust. Timings of methods were measured for different steps along the executions, including preprocessing, postprocessing, each of the different metrics, and the method itself.
### Effect of prior information
To assess the effect of prior information on the performance of a method, we compared the performance of methods which do or do not require the prior information using a two-tailed Mann–Whitney U test. P-values were controlled for multiple testing using Benjamini-Hochberg correction.
### Effect of algorithm components
To assess the effect of algorithm components on the performance of a method, we used (1) a two-tailed Mann–Whitney U test, as implemented in the `wilcox.test` R function, and (2) increase in node purity importance scores using random forest classification, as implemented in the R randomForest package, predicting whether a method scored better than the median score. To evaluate whether a method increased or decreased performance, we used the estimate of the location parameter of the Mann–Whitney U test. We only investigated algorithm components which were part of at least 4 different methods in our evaluation study. P-values were controlled for multiple testing using Benjamini-Hochberg correction.
→
## Scalability[b]
To assess the scalability of each method, we started from 5 real datasets, selected using the centers from a k-medoids as discussed in the synthetic data generation section. We up or down-scaled these datasets … OKAY I GIVE UP generate_dataset.R

>>>>

In this study, we evaluated a total of →`r n_methods_evaluated`← methods in terms of cellular →ordering, topology, scalability, and user friendliness←. Our results indicate that existing methods are typically better at ordering the cells rather than inferring the correct trajectory topology, and that there is no "one-fits-all" method. Instead, the choice toof method →mostly← depends on the trajectory topology present in the data. →To aid users with selecting the most appropriate method, we therefore developed a set of guidelines, available as an interactive app at `r print_url("http://guidelines.dynverse.org")`.← While the field certainly has matured significantly since its inception, further progress is still necessary in order to cope with increasingly complex and novel use cases. We hope to spearhead such developments and the field as a whole, by making our evaluation pipeline easily extensible and freely available at `r print_url("https://www.github.com/dynverse/dynverse")`.

@@

<!-- §2: Plethora of new TI methods -->
A plethora of TI methods has been developed over the last years, and even more are being created every month (`r ref("stable", "tools")`). →Indeed, in several repositories listing single-cell tools,  `r print_url("www.omictools.org")` [@henry_omictoolsinformativedirectory_2014], the "awesome-single-cell" list (`r print_url("https://github.com/seandavi/awesome-single-cell")`) [@davis_awesomesinglecelllistsoftware_2018] and `r print_url("www.scRNA-tools.org")` [@zappia_exploringsinglecellrnaseq_2017], TI methods are one of the largest categories.← All methods have their own unique set of characteristics in terms of underlying algorithm, the required prior information and the produced outputs. Two of the most distinctive differences between TI methods are →whether they fix the topology of the trajectory, and what type(s) of graph topologies can be detected←. Early TI methods typically fixed the topology algorithmically (such as linear [@bendall_singlecelltrajectorydetection_2014; @schlitzer_identificationcdc1cdc2committed_2015; @shin_singlecellrnaseqwaterfall_2015; @campbell_bayesiangaussianprocess_2015] or bifurcating [@haghverdi_diffusionpseudotimerobustly_2016; @setty_wishboneidentifiesbifurcating_2016]), or through parameters provided by the user [@trapnell_dynamics_2014; @matsumoto_scoupprobabilisticmodel_2016]. These methods therefore mainly focused on correctly ordering the cells along the fixed topology. →Most recent methods also infer the topology [@street_slingshotcelllineage_2017; @wolf_graphabstractionreconciles_2017], which increases the difficulty of the problem at hand, but allows an unbiased identification of both the ordering inside a branch and the topology connecting these branches.←

@@

# Results
## Trajectory inference methods
<!-- Method characterisation -->
In this study, we performed a comprehensive evaluation of  `r n_methods_evaluated` TI methods (`r ref("fig", "ti_evaluation_overview", "a")` and `r ref("sfig", "overview_evaluation_expanded")`). In order to make the outputs from different methods directly comparable to each other, we developed a common probabilistic model for representing trajectories from all possible sources (`r ref("fig", "ti_evaluation_overview", "b")`). In this model, the overall topology is represented by a network of "milestones", and the cells are placed within the space formed by each set of connected milestones. Although almost every method returned a unique set of outputs, we were able to classify these outputs into 7 distinct groups, and wrote a common output convertor for each of these groups (`r ref("fig", "results_summary", "a")` and `r ref("sfig", "overview_wrapping")`). When required, we also provided prior information to the method, which ranges from "soft" priors, such as start or end cells, to "hard" priors such as the known grouping of the cells (`r ref("fig", "results_summary", "a")`).

<!-- Topology characterisation -->
The main difference between trajectory inference methods is whether the method fixes the topology, and if it is free, what kind of topologies it can detect. We defined nine possible types of topology, ranging from the very basic (linear, cyclical and bifurcating) to the most complex (connected and disconnected graphs). Most methods either focus on inferring linear trajectories, or limit the search to tree or less complex topologies, with only a selected few attempting to infer cyclic or disconnected topologies (`r ref("fig", "results_summary", "a")`).

<!-- Overall evaluation results -->
We evaluated each method on four core aspects: (i) benchmarking given a gold or silver standard on `r n_datasets_real` real and `r n_datasets_synthetic` synthetic datasets, (ii) their scalability with both the number of cells and features (iii) the stability of the predictions across multiple bootstrap replicates, and (iv) the user friendliness of the provided software and documentation. Overall, we found a large diversity across all four evaluation criteria, with only a few methods, such as PAGA, GNG, Slingshot, Projected TSCAN and SCORPIUS, performing well across the board  (`r ref("fig", "results_summary", "b")`). We will first discuss each evaluation criterion in detail, after which we conclude with guidelines for method users and future perspectives for method developers.
```{r}
add_fig(
  fig_path = result_file("results_summary.pdf", experiment_id = "08-summary"),
  ref_id = "results_summary", 
  caption_main = glue::glue("A characterisation of the {n_methods_evaluated} methods evaluated in this study, and their overall evaluation results."), 
  caption_text = "(a) We characterised the methods according to the wrapper type, their required priors (soft = start or end cells, hard = cell grouping or time course), whether the inferred topology is constrained by the algorithm or a parameter and the types of inferreable topologies. The methods are grouped vertically based on the most complex trajectory type they can infer. (b) Overall results of the evaluation on four criteria: the benchmarking using a reference trajectory on real and synthetic data, the scalability with increasing number of cells and features, the stability across dataset bootstraps, and the implementation quality. ",
  width = 15,
  height = 18
)
```

```{r}
add_fig(
  fig_path = result_file("results_detailed.pdf", experiment_id = "08-summary"),
  ref_id = "results_detailed", 
  caption_main = glue::glue("Detailed results of the four main evaluation criteria: benchmarking, scalability, stability and quality control."), 
  caption_text = "(a) Results of the benchmark across metrics, dataset sources and dataset trajectory types. The performance of a method is generally more stable across dataset sources, but very variable depending on the metric and trajectory type. Also shown is how often a method errored, and what the reasons were for these errors. (b) Scalability results, with increasing number of cells and features. Shown are predicted times at a fixed set of cells (10.000) or features (10.000). (c) Stability results (d) Quality control results across six categories.",
  width = 15,
  height = 18
)
```
## Benchmark
<!-- Benchmark methods -->
We defined several possible metrics to compare a prediction to a reference trajectory (`r ref("snote", "metrics")`). Based on an analysis of their robustness and conformity to a set of rules (`r ref("snote", "metrics")`), we chose 4 metrics each assessing a different aspect of a trajectory: the topology (`r label_metric("him")`), the quality of clustering of the branch assignment  (`r label_metric("F1_branches")`), the cell positions (`r label_metric("correlation")`) and the quality of differentially expressed features  (`r label_metric("featureimp_wcor")`)(`r ref("fig", "ti_evaluation_overview", "d")`).The data compendium consisted of both synthetic datasets, which offer the most exact reference trajectory, and real datasets, which offer the highest biological relevance. These real datasets come from a variety of single-cell technologies, organisms, and dynamic processes, and contain several types of trajectory topologies (`r ref("stable", "datasets")`).  Some real datasets were classified as gold if the reference trajectory was not extracted from the expression data itself, such as via cellular sorting or a time course experiment. All other real datasets were classified as silver. For synthetic datasets we used several data simulators, including our own which simulates gene regulatory networks using a thermodynamic model of gene regulation [@schaffter_genenetweaversilicobenchmark_2011]. For each simulation, we used a real dataset as a reference, to match its dimensions, number of differentially expressed genes, drop-out rates and other statistical properties.

<!-- Benchmark overview -->
We found that method performance was very variable across datasets, indicating that there is no "one-size-fits-all" method that works well on every dataset (`r ref("sfig", "benchmark_interpretation", "a")`). Even methods which can detect most of the trajectory types, such as PAGA or GNG, were not the best methods across all trajectory types (`r ref("fig", "results_detailed", "a")`). The average performance between the different dataset sources was moderately to highly similar correlated, with the performance on all dataset sources being moderately to highly correlated (0.5 - 0.9) with the one on real gold datasets (`r ref("sfig", "benchmark_interpretation", "b")`). On the other hand, the different metrics frequently disagreed with each other, with Monocle scoring better on the topology scores, while other methods, such as Slingshot, were better at ordering the cells and placing them into the correct branches (`r ref("fig", "results_detailed", "a")`).

<!-- Topologies-->
The performance of a method was strongly dependent on the type of trajectory present in the data (`r ref("fig", "results_detailed", "a")`). Slingshot typically performed better on datasets containing more simple topologies, while PAGA, pCreode and RaceID/StemID had higher scores on datasets with trees or more complex trajectories. This was reflected in the types of topologies detected by every method, as those predicted by Slingshot tended to contain less branches, while those detected by PAGA, pCreode and Monocle DDRTree gravitated towards more complex topologies (`r ref("sfig", "benchmark_interpretation", "c")`).

<!-- Complementarity between methods -->
The high variability between datasets, together with the diversity in detected topologies between methods, could indicate some complementarity between the different methods. To test this, we looked for subsets of methods at which percentage of datasets with a "top" model (defined as having a score at least 95% as the best model) was highest. On data which contains a tree trajectory, the use of the best method, RaceID / StemID, would result in getting top model about 50% of the time. This rapidly increased to up to 90% with the addition of 2 to 5 other methods (`r ref("fig", "complementarity", "a")`). We found similar indications of complementarity between the top methods on data containing only linear, bifurcation or multifurcating trajectories (`r ref("fig", "complementarity", "b")`). When the datasets from different trajectory types were pooled, the number of methods necessary to achieve more than 90% top models increased to more than 8, indicating that more methods are necessary when different trajectory types could be present in the data. This pooling also resulted in a more diverse set of methods, containing both strictly linear methods such as SCORPIUS, and methods with a broad trajectory type range such as PAGA. Altogether, this shows that there is considerably complementarity between the different methods, and that users should try out a diverse set of methods on their data.

```{r echo=FALSE, results="asis"}
add_fig(
  fig_path = result_file("complementarity.rds", experiment_id = "10-benchmark_interpretation"),
  ref_id = "complementarity", 
  caption_main = "Complementarity between different TI methods", 
  caption_text = "(a) Here we tested how many methods are needed to still obtain a top model (defined as a model with overall score at least 95\\% of the best model) on datasets containing a tree topology. Shown are the percentage of datasets fulfilling this rule, with increasing number of methods. At each step, we test which method would be the most optimal (as indicated), and chose this method for all next steps. (b) Similar as (a), but now for datasets which contain varying trajectory types.",
  width = 12,
  height = 6
)
```

```{r echo=FALSE, results="asis"}
add_sfig(
  fig_path = result_file("benchmark_interpretation.rds", experiment_id = "10-benchmark_interpretation"),
  ref_id = "benchmark_interpretation", 
  caption_main = "Benchmarking TI methods", 
  caption_text = "Shown are the distributions of the difference in size between predicted and reference topologies. A positive difference means that the topology predicted by the method is more complex than the one in the reference.",
  width = 12,
  height = 15
)
```

## Scalability
While early TI methods were developed at a time where profiling more than a thousand cells was exceptional, methods now have to cope with hundreds of thousands of cells, and perhaps soon with more than ten million [@svensson_exponentialscalingsinglecell_2018]. Moreover, the recent application of TI methods on multi-omics single-cell data also showcases the increasing demands on the number of features [@cao_jointprofilingchromatin_2018]. To assess the scalability, we ran each method on up- and downscaled versions of five distinct real datasets. We then modelled the running time using a generalized additive model (`r ref("sfig", "scaling", "a")`), and assessed whether the method had sublinear, linear or superlinear time complexity with respect to the number of cells and features (`r ref("sfig", "scaling", "b")`). As a control, we compared the predicted time with the actual time on all benchmarking datasets, and found that these were moderately to highly correlated (0.5-0.9) for almost every method.

Overall, methods which detect more complex trajectory types also have longer running times, although two exceptions which stand out from the rest: GNG and PAGA (and its derivatives Projected PAGA and PAGA tree). Despite their ability to detect all trajectory types, these two methods are still able to accurately detect a trajectory with 100.000 cells and 10.000 features in less than 10 minutes (`r ref("fig", "results_detailed", "b")`). Methods with a low running time typically have two defining aspects: they have a sublinear time complexity with respect to the features and/or cells, and adding new cells or features leads to a relatively low increase in time (`r ref("sfig", "scaling", "b")`). We found that over half of all methods had a linear or superlinear complexity with respect to the number of cells, which makes it difficult to apply any of these methods in a reasonable time frame on datasets with more than a thousand of cells (`r ref("sfig", "scaling", "b")`).

```{r echo=FALSE, results="asis"}
add_sfig(
  fig_path = result_file("scaling.pdf", experiment_id = "05-scaling"),
  ref_id = "scaling", 
  caption_main = "Scalability of TI methods", 
  caption_text = "(a) Three examples of average observed running times across five datasets (left) and the predicted running time based on a generalized additive model (GAM, right). (b) Overview of the scalability results of all methods, ordered by their average predicted running time from (a). We used to GAM to predict running times for each method with increasing number of features or cells, and used these values to classify each method into constant / low slope, sublinear, linear and superlinear based on the shape of the curve. We also used the predicted timings to calculate the average increase per new 100 cells or features. Also shown is the average maximal memory usage..",
  width = 12,
  height = 15
)
```
## Stability
←
## Quality control

@@

We observed no clear relation between method quality and method performance (`r ref("fig", "results_summary")`).

@@

<!-- FOR METHOD USERS -->
<!-- Practical guidelines -->
Based on the results of our benchmark, we propose a set of practical guidelines for method users (`r ref("fig", "user_guidelines")` and `r print_url("http://guidelines.dynverse.org")`). We postulate that, as a method's performance is heavily dependent on the trajectory type being studied, the choice of method should be primarily driven by the prior knowledge of the user about which trajectory topology is expected in the data. For the majority of use cases, the user will know very little about the expected trajectory, except perhaps whether the data is expected to contain multiple disconnected trajectories, cycles or a complex tree structure. In each of these use cases, →our evaluation suggests a different set of optimal methods, as shown in `r ref("fig", "user_guidelines")`. Several other factors will also impact the choice of methods, such as the dimensions of the dataset, and the prior information which is available. These factors and several others can all be changed in our interactive app (`r print_url("http://dynverse.org")`). This app can also be used to explore the results of this evaluation, such as filtering the datasets, or changing the importance of the evaluation metrics for the final ranking.←

@@

### Score aggregation
To rank methods, we needed to aggregate the different scores on two levels: across datasets and across different metrics. This aggregation strategy is explained into more detail in `r ref("snote", "metrics")`.

@@

After normalisation, we aggregated step by step the scores from different datasets. We first aggregated the datasets with the same dataset source and trajectory type using an arithmetic mean of their scores. Next, the scores were averaged over different dataset sources, using a arithmetic mean which was weighted based on how much the synthetic and silver scores correlated with the real gold scores. Finally, the scores were aggregated over the different trajectory types again using a arithmetic mean.

Finally, to get an overall benchmarking score, we aggregated the different metrics using a geometric mean.
←
## Method execution
Each execution of a method on a dataset was performed in a separate task as part of a gridengine job. Each task was allocated one CPU core of an Intel(R) Xeon(R) CPU E5-2665 @ 2.40GHz, and one R session was started for each task. During the execution of a method on a dataset, if the time limit (>6h) or memory limit (32GB) was exceeded, or an error was produced, a zero score was returned for that execution. If a method consistently generated errors on this dataset across all replicates, the error is called “data-specific”, otherwise “stochastic”. The environment variable `R_MAX_NUM_DLLS` set to 500. The `base::set.seed` was overridden in order to prevent stochastic TI methods from pretending to be deterministic and/or robust. Timings of methods were measured for different steps along the executions, including preprocessing, postprocessing, each of the different metrics, and the method itself.
→
## Scalability[b]
To assess the scalability of each method, we started from 5 real datasets, selected using the centers from a k-medoids as discussed in the synthetic data generation section. We up or down-scaled these datasets … OKAY I GIVE UP generate_dataset.R
